Ubiquitous Digital Technologies and Spatial Structure; an update

### 2021-03-23 paper/2_final_submission/tranos_ioannides.tex
editing changes

Urban economics suggests that a key source of agglomeration
externalities is proximity, which facilitates interaction and knowledge
spillovers {[}13{]}. Hence, face-to-face interactions and the implied
knowledge spillovers within cities flourish because of lower
transportation and interaction costs {[}11{]}. However, ICT have the
capacity to directly affect this process by further reducing
transportation cost\footnote{Or, more broader, spatial transmission
  costs to follow McCann's terminology {[}14{]}} {[}14{]}. In essence,
the internet and digital communications promote further learning and
matching, which are at the heart of the micro-foundations of
agglomeration economies identified by Duranton and Puga {[}15{]}.
Web-based applications such as Massive Open Online Courses decrease the
need for co-location of actors in order to participate in formal
learning activities. Furthermore, online social media such as
LinkedIn\footnote{LinkedIn has been identified as ``the largest
  professional matchmaker site in the world'' {[}16{]}, p.~207. See also
  {[}17{]}.} can enhance the probability of matching and indeed the
quality of matches even within cities. Therefore, it is natural to ask
whether the widespread adoption of ICT could offset the benefits of
immediate physical proximity and result in more dispersed spatial
structures, or further reinforce agglomeration externalities and lead to
more concentrated spatial structures.

@@

--\textgreater{} Kolko {[}22{]} uses internet diffusion data and
identifies a clear complementary link between internet usage and city
size. Interestingly, he identifies higher internet domain densities in
remote cities which indicates a substitution effect of the internet for
longer-distance non-electronic communications. His results are
consistent for different measures of internet diffusion (internet domain
density and internet take-up). Sinai and Waldfogel {[}23{]} approach the
same question from the consumers' point of view and find support for
complementarity between internet and cities. They study the link between
market size and locally targeted online content and find more online
local content in larger markets. However, they also obtain evidence for
a substitution effect: holding local online content constant, market
size has a negative effect on individual connectivity. Forman et
al.~{[}24{]} examines whether commercial internet adoption is higher in
cities than in rural areas. While the former would indicate a
complementarity between internet adoption and cities, the latter would
reflect a substitutability relation, according to which the internet is
used as a means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this is not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

--\textgreater{} Focusing on rural areas, Partridge et al.~{[}28{]} find
no evidence that rural distance penalties in the US have substantially
changed since \(1970s\) indicating that technological changes including
the internet and digital communications have not affected spatial
structure. Also interesting are the findings of a more recent study by
Kim and Orazem {[}29{]} on the economic effects of broadband internet in
rural areas. They identify a positive effect on new firm location
decisions, with the effect being higher in rural areas with larger
population and in rural areas adjacent to a metropolitan area. These
results suggest a complementary relation between the internet and
agglomeration economies.

--\textgreater{} Moving beyond modelling the direct relationship between
ICT and spatial structure, Bekkerman and Gilpin {[}30{]} focused on the
role of locally based information resources using a dataset about the US
libraries during the period \(2000-2008\). Their results suggest that
internet access increases the demand and the value of locally accessible
information, and such complementarities are higher in larger
metropolitan areas. Anenberg and Kung {[}31{]} also identified a
complementary relation between the internet and consumption variety in
cities by focusing on food truck industry in the US.

\color{black}

Thus, clearly, previous research suggests that the declaration of the
``death of distance'' has proven to be premature {[}32{]}. However, the
exact impact of digital communications on spatial structure is still an
open question. As Leamer and Storper {[}33{]} indicate, the internet can
affect both centripetal and centrifugal forces. The only cross country
study {[}18{]} supports a clear substitution effect, whereas most of the
above studies have focused either on the US or on some specific cities.
Moreover, most of the above studies examined the
complementarity/substitutability question over time periods when the
internet and other digital communication technologies were still
emerging. For instance, internet penetration in the US in \(2000\),
which was the focus for quite a few of the above studies including
Ioannides et al.~{[}18{]}, was just above \(50\) percent, while in
\(2016\) it reached almost \(90\) percent {[}34{]}. At a global scale
internet penetration increased from \(7\) to \(46\) percent during the
same period {[}35{]}. In addition, although email and instant messaging
technologies were widespread in the developed world in early \(2000s\),
network externalities due to mobile internet and online social media
were nowhere close to what we are familiar with today. For instance,
Facebook users increased from \(1\) million in \(2004\) to more than
\(1.5\) billion in \(2015\)\footnote{\url{http://newsroom.fb.com/company-info/}}.
Hence, it might have been premature for the spatial economic effects of
ICT adoption to have been materialized by the time that most of the
above studies were conducted.

@@

The present paper does not address the issue that adoption of ICT may
have different effects on urban dispersion on a national scale, that is
how far are major urban areas, that is, cities, from one another, versus
dispersion within major urban areas, that is urban sprawl. Such an
inquiry would require detailed data about patterns of how urban sprawl
may hamper or promote economic interaction, given advancements in
information technology. We think that data from the Covid-19 pandemic
would likely throw light at this issue in the future, as it is clear
that not all economic activities might continue in an unencumbered
manner; see Dingel and Neiman {[}36{]}. \color{black}

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Pareto
exponent \color{black} for a broad sample of countries over time. The
Pareto exponent \color{black} is one of the most widely used measures of
spatial dispersion with numerous applications in urban economics and
economic geography ({[}see for example {[}37{]}; {[}38{]}; {[}39{]};
{[}40{]}; {[}41{]}; {[}42{]}; {[}43{]}). This an appropriate measure of
dispersion because of the extreme heterogeneity of the city size
distribution and the very good fits normally obtained with such
estimations\footnote{We also replicate our analysis using other measures
  of dispersion in Appendix S1. Our results remain consistent.}.
\color{black} Skipping details pertaining of the suitability of the
approach, which may be found in Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, we proceed with the estimation of the logarithmic
form of:

@@

This leads to an estimate of \(\zeta\), known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere
to for the remained of the text, even when \(\zeta\) is estimated to be
near \(1\). The term Zipf coefficient should be reserved for the special
case of \(\zeta=1\). While much of the literature focuses on the Pareto
exponent being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\): \color{black}

@@

The estimation of equation \ref{eq2} has typically been performed by
Ordinary Least Squares (OLS). Gabaix and Ioannides {[}44{]} discuss the
downward bias of estimates of equation \ref{eq2} using OLS on small
samples. Gabaix and Ibragimov {[}45{]} propose a practical remedy to
correct this bias, which we do adopt in this paper: instead of using the
\(log\) of \(rank\) of a city \(i\) in a country \(c\) in year \(t\),
they propose to use the \(log(rank-0.5)\), which has indeed been widely
adopted. Researchers working in this area must contend with definitional
differences as well as differences in availability of different kinds of
data sources. Definitions of cities differ across countries, for
political, administrative and legal reasons\footnote{See {[}46{]}, Ch.
  8, for issues and pitfalls associated with different definitions.}. In
contrast to the data drawn from Thomas Brinkhoff's City Population
project {[}47{]}\footnote{\url{http://www.citypopulation.de}}, which
were used by Soo {[}48{]}, Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, the multi-country analysis in the present paper is
based on the annual population data for urban agglomerations with
\(300,000\) inhabitants or more from the Department of Economic and
Social Affairs of the United Nations {[}9{]}. Despite some criticism
about the consistency of the urban agglomeration definitions across
different countries {[}49{]}, this is the only available source for
yearly, multi-country population data for urban agglomerations
{[}50--52{]}. The results based on these data are reported and discussed
in \protect\hyperlink{sec4}{Section 4}.

@@

\color{black}

\noindent The model enable us to estimate the effects of a vector \(X\)
of explanatory variables which are included to account for the spatial
structure of country \(c\) in year \(t\). The main variables of interest
here are internet and digital communications variables that include:
internet users per \(100\) inhabitants, broadband users per \(100\)
inhabitants, mobile phone users per \(100\) inhabitants, and fixed phone
users per \(100\) inhabitants. To address a potential omitted variable
bias equation \ref{eq3} includes country (\(\theta_c\)) and year
(\(\delta_t\)) fixed effects; \color{black} \(e_{ct}\) is the error
term. In addition, vector \(X\) includes several control variables,
whose descriptive statistics together with those for the other variables
used to estimate \ref{eq3} are reported in Table \ref{desc.global}.

@@

As Table \ref{cor} indicates, although there some rather strong
correlations among these variables, mobile phone penetration appears to
have a distinct character from its fixed phone counterpart: their
correlation coefficient is only \(0.445\). This probably highlights the
different composition of the population or infrastructure development
patterns in the developing world, where mobile telephony helped overcome
the lack of fixed line infrastructure and mobile phone networks are also
used as the main way to use the internet {[}53,54{]}.

@@

In order to capture the micro-dynamics of the urban systems in the two
case studies, we follow Batty {[}55{]} and Havlin {[}56{]} and focus on
the difference in ranks for individual cities during the study
period.\footnote{Changes in ranking of individuals has also been used
  elsewhere in economics because of its robustness properties. C.f.
  Chetty and Hendren {[}57{]}.} Contrary to their approach, we are
interested in the real difference in rankings rather than in absolute
differences, in order to capture whether or a city improves its ranking
within the urban system during the study period. We then test whether
our internet-related variable has an effect on the change in ranking.
Hence, we define the left-hand side (LHS) variable of equation \ref{eq4}
as follows:

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Pareto exponent, \color{black} as estimated
according to the Gabaix and Ibragimov {[}45{]} correction. The main
explanatory variables of interest are provided at the top of Table
\ref{ols.global}; namely internet, broadband, mobile and fixed telephony
per \(100\) habitants, expressed in natural logarithms. The main
variables of interest are introduced successively on their own in the
regressions reported in Table \ref{ols.global}. All regressions include
country fixed effects to control for unobserved heterogeneity and a time
trend. In addition, the observations are weighted with the inverse
squared standard error of the estimated Pareto exponent \color{black}
(see Table \ref{desc.global}) to address potential noise that is carried
over from the first part of our identification strategy. In regards to
the interpretation of the estimated coefficients, given that the Pareto
exponent \color{black} has entered the regression not as an absolute
value, but instead as a real number a positive coefficient for a RHS
variable indicates an impact towards the decrease of the spatial
dispersion of population. In other words, a positive coefficient
indicates an effect towards less uniform city sizes that is more
dispersion of city sizes. The latter is indicative of enhancement of
agglomeration economies because of the expansion of digital
technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Pareto exponent, \color{black} might be affected by
another source, which also affects internet penetration. For instance,
economic development might affect the concentration of population in
large cities and at the same time enable more people to go online. If we
do not address this issue, the coefficient for the main variable of
interest will capture potential effects that internet penetration has on
spatial structure, but also potential reverse causality effects that
spatial structure might generate on internet penetration. To overcome
this potential problem, Table \ref{2sls.global} reports estimates of
equation \ref{eq3} using two-stage least squares (2SLS) with
instrumental variables (IVs). The latter are variables which are
correlated with our endogenous variables, but do not influence current
spatial structure. Such an approach will enable us to estimate the
causal effect -- if any -- of the internet and digital communications on
spatial structure. At a first stage, our endogenous variables are
regressed against the IVs. Then, the predicted values of the endogenous
variables based on the IVs and the other control variables are used
instead of the endogenous variable to estimate eq. \ref{eq3}. A
significant effect will verify the causal impact of the internet and
digital communication usage on spatial structure.

@@

The main challenge for such an exercise is to find a valid IV. In our
case, the challenge is even bigger as we need to find a variable which
is not only fit for purpose, but it also contain as few missing values
as possible in order to retain the same number of observations as the
OLS models. This is particularly difficult for a diverse sample of
countries as the one we are dealing with here. We propose here the
percentage of female participation in the labor force, which is directly
related to digital infrastructure, but not to spatial structure (see
Table \ref{desc.global} for the descriptive statistics). The literature
provides empirical and theoretical evidence that ICT are correlated with
female participation in labour markets mostly because they enable
teleworking {[}58--62{]} \color{black} At the same time, female labor
participation is not related with spatial structure especially given the
diverse sample of countries. Although one might think that increased
female labor participation may lead to relocation of households to large
cities and therefore affect the spatial structure of cities, such a
process varies a lot within our dataset. Even if this might be true for
developed countries with mature urban systems for which a location
within a large urban center might provide opportunities for both male
and female workers to find jobs, our dataset also includes countries,
\(41\) per cent of the GDP of which is attributed to primary activities
(see Table \ref{desc.global}). So, female labor participation might also
be related to economic activities located outside large urban centers.
In terms of the relevant tests presented in Table \ref{2sls.global} the
weak identification first stage F-test exceeds by far all the Stock-Yogo
critical values for all, but the mobile telephony regression. Hence, we
are confident to interpret the effects of internet, broadband internet
and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that
increases in internet and broadband internet usage have resulted to a
decrease of the spatial dispersion of population for the time period and
for the panel of countries included in our data. In other words,
increases in internet and digital communications resulted to national
urban systems which are less uniform in terms of city sizes and are
characterized by higher population concentrations in larger cities. The
exact opposite appears to happen for fixed telephony, which our results
indicate that has led to an increase of spatial dispersion of
population. We expect that the IV estimates capture the local average
treatment effect {[}63{]} of these labour markets, which are
characterised by more active engagement with the internet, and,
therefore, experience a larger effect on the Pareto exponents.
\color{black} Importantly, our identification strategy enables us to
address potential reverse causality issues and treat the results of
Table \ref{2sls.global} as causal. Hence, the main finding of our
multi-country, global analysis based on data for urban agglomerations of
at least 300,000 inhabitants is that internet and broadband internet
usage appear to act in favor of agglomeration economies and result to
urban systems with more dominant cities on the top of the urban
hierarchy. On the contrary, fixed telephony usage acted in favour of
centrifugal forces. Our identification strategy does not enable us to
make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of the Pareto exponent \color{black} we only included
urban agglomeration of \(300,000\) inhabitants or more due to data
availability. Hence, the above estimations cannot verify whether such an
effect is also valid for smaller cities. Interestingly, the above
results regarding the role of fixed telephony are in agreement with
previous results from {[}18{]} {[}19{]} which used a smaller panel of
countries but without constraints regarding the city size. However, our
results do not support the findings of {[}19{]} regarding internet
usage.

It is indeed puzzling that the results on Table \ref{2sls.global} that
the effects on urban concentration for internet users and broadband
users, on the one hand, and mobile and fixed telephony users are in
opposite directions. We wish to broach this subject in the context of
some related literature, namely {[}64{]} and {[}65{]}. The latter
supports the notion that ICT favors agglomeration in the sense that in
spite of their convenience modern technologies do no make up for the
need of proximity in social interactions. The former supports the notion
people who communicate more frequently are likely to be near one
another. ICT, as viewed through such granular studies of individuals'
communications do not make up for physical distance; individuals nearer
one another communicate more.\\
\color{black}

@@

\leavevmode\hypertarget{ref-dingel2020many}{}%
36. Dingel JI, Neiman B. How many jobs can be done at home? National
Bureau of Economic Research; 2020.
doi:\href{https://doi.org/DOI\%2010.3386/w26948}{DOI 10.3386/w26948}

\leavevmode\hypertarget{ref-black_urban_2003}{}%
37. Black D, Henderson V. Urban evolution in the USA. Journal of
economic geography. 2003;3: 343--372.

>>>>

Urban economics suggests that a key source of agglomeration
externalities is proximity, which facilitates interaction and knowledge
spillovers {[}13{]}. Hence, face-to-face interactions and the implied
knowledge spillovers within cities flourish because of lower
transportation and interaction costs {[}11{]}. However, ICT have the
capacity to directly affect this process by further reducing
transportation cost {[}14{]}. In essence, the internet and digital
communications promote further learning and matching, which are at the
heart of the micro-foundations of agglomeration economies identified by
Duranton and Puga {[}15{]}. Web-based applications such as Massive Open
Online Courses decrease the need for co-location of actors in order to
participate in formal learning activities. Furthermore, online social
media such as LinkedIn can enhance the probability of matching and
indeed the quality of matches even within cities. This is why LinkedIn
has been identified as ``the largest professional matchmaker site in the
world'' {[}{[}16{]} p.~207; {[}17{]}{]}. Therefore, it is natural to ask
whether the widespread adoption of ICT could offset the benefits of
immediate physical proximity and result in more dispersed spatial
structures, or further reinforce agglomeration externalities and lead to
more concentrated spatial structures.

@@

Kolko {[}22{]} uses internet diffusion data and identifies a clear
complementary link between internet usage and city size. Interestingly,
he identifies higher internet domain densities in remote cities which
indicates a substitution effect of the internet for longer-distance
non-electronic communications. His results are consistent for different
measures of internet diffusion (internet domain density and internet
take-up). Sinai and Waldfogel {[}23{]} approach the same question from
the consumers' point of view and find support for complementarity
between internet and cities. They study the link between market size and
locally targeted online content and find more online local content in
larger markets. However, they also obtain evidence for a substitution
effect: holding local online content constant, market size has a
negative effect on individual connectivity. Forman et al.~{[}24{]}
examines whether commercial internet adoption is higher in cities than
in rural areas. While the former would indicate a complementarity
between internet adoption and cities, the latter would reflect a
substitutability relation, according to which the internet is used as a
means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this is not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

Focusing on rural areas, Partridge et al.~{[}28{]} find no evidence that
rural distance penalties in the US have substantially changed since
\(1970s\) indicating that technological changes including the internet
and digital communications have not affected spatial structure. Also
interesting are the findings of a more recent study by Kim and Orazem
{[}29{]} on the economic effects of broadband internet in rural areas.
They identify a positive effect on new firm location decisions, with the
effect being higher in rural areas with larger population and in rural
areas adjacent to a metropolitan area. These results suggest a
complementary relation between the internet and agglomeration economies.

Moving beyond modelling the direct relationship between ICT and spatial
structure, Bekkerman and Gilpin {[}30{]} focused on the role of locally
based information resources using a dataset about the US libraries
during the period \(2000-2008\). Their results suggest that internet
access increases the demand and the value of locally accessible
information, and such complementarities are higher in larger
metropolitan areas. Anenberg and Kung {[}31{]} also identified a
complementary relation between the internet and consumption variety in
cities by focusing on food truck industry in the US.

Thus, clearly, previous research suggests that the declaration of the
``death of distance'' has proven to be premature {[}32{]}. However, the
exact impact of digital communications on spatial structure is still an
open question. As Leamer and Storper {[}33{]} indicate, the internet can
affect both centripetal and centrifugal forces. The only cross country
study {[}18{]} supports a clear substitution effect, whereas most of the
above studies have focused either on the US or on some specific cities.
Moreover, most of the above studies examined the
complementarity/substitutability question over time periods when the
internet and other digital communication technologies were still
emerging. For instance, internet penetration in the US in \(2000\),
which was the focus for quite a few of the above studies including
Ioannides et al.~{[}18{]}, was just above \(50\) percent, while in
\(2016\) it reached almost \(90\) percent {[}34{]}. At a global scale
internet penetration increased from \(7\) to \(46\) percent during the
same period {[}35{]}. In addition, although email and instant messaging
technologies were widespread in the developed world in early \(2000s\),
network externalities due to mobile internet and online social media
were nowhere close to what we are familiar with today. For instance,
Facebook users increased from \(1\) million in \(2004\) to more than
\(1.5\) billion in \(2015\) {[}36{]}. Hence, it might have been
premature for the spatial economic effects of ICT adoption to have been
materialized by the time that most of the above studies were conducted.

@@

The present paper does not address the issue that adoption of ICT may
have different effects on urban dispersion on a national scale, that is
how far are major urban areas, that is, cities, from one another, versus
dispersion within major urban areas, that is urban sprawl. Such an
inquiry would require detailed data about patterns of how urban sprawl
may hamper or promote economic interaction, given advancements in
information technology. We think that data from the Covid-19 pandemic
would likely throw light at this issue in the future, as it is clear
that not all economic activities might continue in an unencumbered
manner; see Dingel and Neiman {[}37{]}.

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Pareto
exponent for a broad sample of countries over time. The Pareto exponent
is one of the most widely used measures of spatial dispersion with
numerous applications in urban economics and economic geography ({[}see
for example {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]}; {[}42{]}; {[}43{]};
{[}44{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations. We also replicate our
analysis using other measures of dispersion in Appendix S1. Our results
remain consistent. Skipping details pertaining of the suitability of the
approach, which may be found in Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, we proceed with the estimation of the logarithmic
form of:

@@

This leads to an estimate of \(\zeta\), known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere
to for the remained of the text, even when \(\zeta\) is estimated to be
near \(1\). The term Zipf coefficient should be reserved for the special
case of \(\zeta=1\). While much of the literature focuses on the Pareto
exponent being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\):

@@

The estimation of equation \ref{eq2} has typically been performed by
Ordinary Least Squares (OLS). Gabaix and Ioannides {[}45{]} discuss the
downward bias of estimates of equation \ref{eq2} using OLS on small
samples. Gabaix and Ibragimov {[}46{]} propose a practical remedy to
correct this bias, which we do adopt in this paper: instead of using the
\(log\) of \(rank\) of a city \(i\) in a country \(c\) in year \(t\),
they propose to use the \(log(rank-0.5)\), which has indeed been widely
adopted. Researchers working in this area must contend with definitional
differences as well as differences in availability of different kinds of
data sources. Definitions of cities differ across countries, for
political, administrative and legal reasons {[}see {[}47{]}, Ch. 8, for
issues and pitfalls associated with different definitions{]}. In
contrast to the data drawn from Thomas Brinkhoff's City Population
project {[}48{]}, which were used by Soo {[}49{]}, Ioannides et
al.~{[}18{]} and Tranos and Ioannides {[}19{]}, the multi-country
analysis in the present paper is based on the annual population data for
urban agglomerations with \(300,000\) inhabitants or more from the
Department of Economic and Social Affairs of the United Nations {[}9{]}.
Despite some criticism about the consistency of the urban agglomeration
definitions across different countries {[}50{]}, this is the only
available source for yearly, multi-country population data for urban
agglomerations {[}51--53{]}. The results based on these data are
reported and discussed in \protect\hyperlink{sec4}{Section 4}.

@@

\noindent The model enable us to estimate the effects of a vector \(X\)
of explanatory variables which are included to account for the spatial
structure of country \(c\) in year \(t\). The main variables of interest
here are internet and digital communications variables that include:
internet users per \(100\) inhabitants, broadband users per \(100\)
inhabitants, mobile phone users per \(100\) inhabitants, and fixed phone
users per \(100\) inhabitants. To address a potential omitted variable
bias equation \ref{eq3} includes country (\(\theta_c\)) and year
(\(\delta_t\)) fixed effects; \(e_{ct}\) is the error term. In addition,
vector \(X\) includes several control variables, whose descriptive
statistics together with those for the other variables used to estimate
\ref{eq3} are reported in Table \ref{desc.global}.

@@

As Table \ref{cor} indicates, although there some rather strong
correlations among these variables, mobile phone penetration appears to
have a distinct character from its fixed phone counterpart: their
correlation coefficient is only \(0.445\). This probably highlights the
different composition of the population or infrastructure development
patterns in the developing world, where mobile telephony helped overcome
the lack of fixed line infrastructure and mobile phone networks are also
used as the main way to use the internet {[}54,55{]}.

@@

In order to capture the micro-dynamics of the urban systems in the two
case studies, we follow Batty {[}56{]} and Havlin {[}57{]} and focus on
the difference in ranks for individual cities during the study period.
Changes in ranking of individuals have also been used elsewhere in
economics because of its robustness properties. C.f. Chetty and Hendren
{[}58{]}. Contrary to their approach, we are interested in the real
difference in rankings rather than in absolute differences, in order to
capture whether or a city improves its ranking within the urban system
during the study period. We then test whether our internet-related
variable has an effect on the change in ranking. Hence, we define the
left-hand side (LHS) variable of equation \ref{eq4} as follows:

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Pareto exponent, as estimated according to
the Gabaix and Ibragimov {[}46{]} correction. The main explanatory
variables of interest are provided at the top of Table \ref{ols.global};
namely internet, broadband, mobile and fixed telephony per \(100\)
habitants, expressed in natural logarithms. The main variables of
interest are introduced successively on their own in the regressions
reported in Table \ref{ols.global}. All regressions include country
fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared
standard error of the estimated Pareto exponent (see Table
\ref{desc.global}) to address potential noise that is carried over from
the first part of our identification strategy. In regards to the
interpretation of the estimated coefficients, given that the Pareto
exponent has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

@@

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Pareto exponent, might be affected by another source,
which also affects internet penetration. For instance, economic
development might affect the concentration of population in large cities
and at the same time enable more people to go online. If we do not
address this issue, the coefficient for the main variable of interest
will capture potential effects that internet penetration has on spatial
structure, but also potential reverse causality effects that spatial
structure might generate on internet penetration. To overcome this
potential problem, Table \ref{2sls.global} reports estimates of equation
\ref{eq3} using two-stage least squares (2SLS) with instrumental
variables (IVs). The latter are variables which are correlated with our
endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any
-- of the internet and digital communications on spatial structure. At a
first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs
and the other control variables are used instead of the endogenous
variable to estimate eq. \ref{eq3}. A significant effect will verify the
causal impact of the internet and digital communication usage on spatial
structure.

@@

The main challenge for such an exercise is to find a valid IV. In our
case, the challenge is even bigger as we need to find a variable which
is not only fit for purpose, but it also contain as few missing values
as possible in order to retain the same number of observations as the
OLS models. This is particularly difficult for a diverse sample of
countries as the one we are dealing with here. We propose here the
percentage of female participation in the labor force, which is directly
related to digital infrastructure, but not to spatial structure (see
Table \ref{desc.global} for the descriptive statistics). The literature
provides empirical and theoretical evidence that ICT are correlated with
female participation in labour markets mostly because they enable
teleworking {[}59--63{]} At the same time, female labor participation is
not related with spatial structure especially given the diverse sample
of countries. Although one might think that increased female labor
participation may lead to relocation of households to large cities and
therefore affect the spatial structure of cities, such a process varies
a lot within our dataset. Even if this might be true for developed
countries with mature urban systems for which a location within a large
urban center might provide opportunities for both male and female
workers to find jobs, our dataset also includes countries, \(41\) per
cent of the GDP of which is attributed to primary activities (see Table
\ref{desc.global}). So, female labor participation might also be related
to economic activities located outside large urban centers. In terms of
the relevant tests presented in Table \ref{2sls.global} the weak
identification first stage F-test exceeds by far all the Stock-Yogo
critical values for all, but the mobile telephony regression. Hence, we
are confident to interpret the effects of internet, broadband internet
and fixed telephony usage.

The estimations presented in Table \ref{2sls.global} indicate that
increases in internet and broadband internet usage have resulted to a
decrease of the spatial dispersion of population for the time period and
for the panel of countries included in our data. In other words,
increases in internet and digital communications resulted to national
urban systems which are less uniform in terms of city sizes and are
characterized by higher population concentrations in larger cities. The
exact opposite appears to happen for fixed telephony, which our results
indicate that has led to an increase of spatial dispersion of
population. We expect that the IV estimates capture the local average
treatment effect {[}64{]} of these labour markets, which are
characterised by more active engagement with the internet, and,
therefore, experience a larger effect on the Pareto exponents.
Importantly, our identification strategy enables us to address potential
reverse causality issues and treat the results of Table
\ref{2sls.global} as causal. Hence, the main finding of our
multi-country, global analysis based on data for urban agglomerations of
at least 300,000 inhabitants is that internet and broadband internet
usage appear to act in favor of agglomeration economies and result to
urban systems with more dominant cities on the top of the urban
hierarchy. On the contrary, fixed telephony usage acted in favour of
centrifugal forces. Our identification strategy does not enable us to
make similar interpretations for mobile telephony usage.

@@

In total, the results indicate that our measures of internet and
telephony penetration have further enhanced agglomeration forces, at
least for large urban agglomerations, while the opposite happens for
fixed telephony. These results are robust against endogeneity issues,
but are limited to urban agglomeration included in our data. Indeed, for
the estimation of the Pareto exponent we only included urban
agglomeration of \(300,000\) inhabitants or more due to data
availability. Hence, the above estimations cannot verify whether such an
effect is also valid for smaller cities. Interestingly, the above
results regarding the role of fixed telephony are in agreement with
previous results from {[}18{]} {[}19{]} which used a smaller panel of
countries but without constraints regarding the city size. However, our
results do not support the findings of {[}19{]} regarding internet
usage.

It is indeed puzzling that the results on Table \ref{2sls.global} that
the effects on urban concentration for internet users and broadband
users, on the one hand, and mobile and fixed telephony users are in
opposite directions. We wish to broach this subject in the context of
some related literature, namely {[}65{]} and {[}66{]}. The latter
supports the notion that ICT favors agglomeration in the sense that in
spite of their convenience modern technologies do no make up for the
need of proximity in social interactions. The former supports the notion
people who communicate more frequently are likely to be near one
another. ICT, as viewed through such granular studies of individuals'
communications do not make up for physical distance; individuals nearer
one another communicate more.

@@

\leavevmode\hypertarget{ref-fb}{}%
36. Facebook. Facebook company information. 2021. Available:
\url{http://newsroom.fb.com/company-info/}

\leavevmode\hypertarget{ref-dingel2020many}{}%
37. Dingel JI, Neiman B. How many jobs can be done at home? National
Bureau of Economic Research; 2020.
doi:\href{https://doi.org/DOI\%2010.3386/w26948}{DOI 10.3386/w26948}

\leavevmode\hypertarget{ref-black_urban_2003}{}%
38. Black D, Henderson V. Urban evolution in the USA. Journal of
economic geography. 2003;3: 343--372.

\leavevmode\hypertarget{ref-frenken_theoretical_2007}{}%
39. Frenken K, Boschma RA. A theoretical framework for evolutionary
economic geography: Industrial dynamics and urban growth as a branching
process. Journal of economic geography. 2007;7: 635--649.
